Python as a machine learning language: experiences from the \textit{scikit-learn} project

### 2013-06-05 paper.tex
Data representation

The \textit{scikit-learn} project~\citep{pedregosa2011} is an open source
machine learning library for the Python programming language. The ambition of
the project is to provide  effective and long-established machine learning tools
within a programming environment that is easy-to-use, efficient, accessible to
non-machine learning experts and reusable in various scientific areas. In
itself, the project is a library of machine learning tools including, among
others, learning algorithms, model selection tools and preprocessing procedures
rather than being a novel domain-specific programming language. From its core,
the library has been designed to tie in with the set of numeric and scientific
packages centered around the Numpy and SciPy~\citep{vanderwalt2011,
varoquaux2013scipy} libraries. As such, \textit {scikit-learn} is intended to be
used within Python scripts and requires some Python programming knowledge. At
the time, it offers no command line or graphical user interface.

@@

\subsection{Data formats}

For a machine learning library, one of the most important decisions we had to
make is how to represent data.  Rather than reinventing the wheel, we opted for
NumPy multidimensional arrays \Citep{vanderwalt2011} for dense data and SciPy
sparse matrices for sparse data.  While those may seem like a ``bare''
representation of data when compared to more object-oriented representations
(Weka is an example of this style of representation, \citealp{hall2009weka}),
it brings the prime advantage of allowing us to rely on NumPy and SciPy's
\textit{vector operations} which are often orders of magnitude faster
than the corresponding Python loops,
while at the same time keeping the code short and readable.
Conversion to NumPy and SciPy formats is usually easy and many scientific users
of Python will already be used to such formats, since they are pervasive in
other scientific Python packages.  For tasks where input is likely to consist
of text files or semi-structured objects, we provide ``vectorizers'' -- objects
that efficiently convert such data to the NumPy or SciPy formats.
% FIXME: elaborate what is meant below
% Lars: elaborated, but I'm not sure how important this is.
% It mostly reflects experiences with OO libraries that require conversions
% between types just to pass information to a method.
%Finally, offering a custom datatype for samples would require a conversion
%\textit{in any case}, since such a datatype would be toolkit-specific
%and no program would store its data using such a type
%if it weren't from the outset designed to use scikit-learn.

>>>>

The \textit{scikit-learn} project~\citep{pedregosa2011} is an open source
machine learning library for the Python programming language. The ambition of
the project is to provide  effective and long-established machine learning tools
within a programming environment that is accessible to non-machine learning
experts, efficient, and reusable in various scientific areas. In itself, the
project is a library of machine learning tools  rather than  a novel domain-
specific programming language. Among other things, the library includes
classical learning algorithms, model selection tools and preprocessing
procedures. From its core, the library has been designed to tie in with the set
of numeric and scientific packages centered around the Numpy and
SciPy~\citep{vanderwalt2011, varoquaux2013scipy} libraries. As such, \textit
{scikit-learn} is intended to be used within Python scripts and requires some
Python programming knowledge. At the time, it offers no command line or
graphical user interface.

@@

\subsection{Data representation}

In most machine learning tasks, data is modeled as a set of variables.  For
example, in a classification or regression task, the goal is to find a mapping
between input variables $X_1, ..., X_p$ -- called attributes or features -- and
some output variable $Y$. A sample is then defined as a tuple of values $(x_1,
..., x_p, y)$ of these variables and a dataset as a collection of such samples.
As such, matrices of numerical values\footnote{In the case of categorical
variables, numerical values can be obtained using various encoding schemes, such
as one-hot encoding.} constitute a natural representation used almost
unanimously within all areas of machine learning. In particular, each row in the
matrix then corresponds to a sample of the dataset and each column to one of the
variables of the problem.

In \textit{scikit-learn}, we chose a representation of data that is as close as
possible to the matrix representation: datasets are encoded as NumPy
multidimensional arrays for dense data and as SciPy sparse matrices for sparse
data. While those may seem like bare representations of data when compared to
more object-oriented representations\footnote{Weka~\citep{hall2009weka} is an
example of a toolkit-specific object-oriented data representation.}, they bring
the prime advantage of allowing us to rely on all NumPy and SciPy vector
operations -- often orders of magnitude faster that the corresponding Python
loops -- while allowing us at the same time to keep the code short and readable.
From a practical point of view, these formats also provide a bunch of data
loading and conversion tools which make them very easy to use in many
contexts\footnote{For tasks where the inputs are text files or semi-structured
objects, we provide \textit{vectorizer} objects that efficiently convert such
data to the NumPy or SciPy formats.}. Finally, this design choice has also been
motivated by the fact that, given their pervasiveness in many other scientific
Python packages, many scientific users of Python are likely to be already used
to those formats, hence helping them to familiarize with the project.

% For a machine learning library, one of the most important decisions we had to
% make is how to represent data.  Rather than reinventing the wheel, we opted for
% NumPy multidimensional arrays \Citep{vanderwalt2011} for dense data and SciPy
% sparse matrices for sparse data.  While those may seem like a ``bare''
% representation of data when compared to more object-oriented representations
% (Weka is an example of this style of representation, \citealp{hall2009weka}),
% it brings the prime advantage of allowing us to rely on NumPy and SciPy's
% \textit{vector operations} which are often orders of magnitude faster
% than the corresponding Python loops,
% while at the same time keeping the code short and readable.
% Conversion to NumPy and SciPy formats is usually easy and many scientific users
% of Python will already be used to such formats, since they are pervasive in
% other scientific Python packages.  For tasks where input is likely to consist
% of text files or semi-structured objects, we provide ``vectorizers'' -- objects
% that efficiently convert such data to the NumPy or SciPy formats.

% FIXME: elaborate what is meant below
% Lars: elaborated, but I'm not sure how important this is.
% It mostly reflects experiences with OO libraries that require conversions
% between types just to pass information to a method.
%Finally, offering a custom datatype for samples would require a conversion
%\textit{in any case}, since such a datatype would be toolkit-specific
%and no program would store its data using such a type
%if it weren't from the outset designed to use scikit-learn.
